\bfseries An Extension of the Sandwich theorem for two-sided limits

### 2025-07-06 lecturenotes_aima203/ch7_aima203.tex
chap 7

\section{Taylor's series method}
We consider the IVP ~\ref{ivp}. If $y(x)$ is the exact solution of this IVP, then the Taylor's series for $y(x)$ around $x=x_0$ is given by
\begin{equation}
  \label{ivptay}
  y(x) = y_0 + (x-x_0)y'_0 + \frac{(x-x_0)^2}{2!}\, y''_0 + ...
\end{equation}
Differentiating ~\ref{ivp} successively, with respect to $x$, we obtain, $y'', y''', ...$. Since ~\ref{ivptay} is an infinite series, for $y(x_0)$ correct to $N$ decimal places we solve the following equation \\
$n^{th}$ term of Taylor series $\leq$ Upper limiting error = $\displaystyle \frac{(x-x_0)^n}{n!} \, f^{(n)} (x_0) = \frac{1}{2} 10^{-N}$. \\[1mm]
If this holds then, the $n$ number of terms is sufficient otherwise, we need to include additional terms.

\section{Picard's Method of Successive Approximation}
Integrating the above differential equation ~\ref{ivp}, we obtain
\begin{equation}
  \label{picard1}
  y = y_0 + \int_{x_0}^x \; f(x,y)\,dx
\end{equation}
This equation in which the unknown function $y$ appears under the integral sign is called an \textit{integral equation}. Such equation can be solved by the method of successive approximations as follows:
\begin{equation}
  \label{picard2}
  y^{(1)} = y_0 + \int_{x_0}^x \; f(x,y_0)\,dx
\end{equation}
Proceeding in this way, we obtain:
\begin{equation}
  \label{picardit}
    y^{(n)} = y_0 + \int_{x_0}^x \; f(x,y^{(n-1)}\,dx \hspace{5mm} \text{with} \hspace{3mm}  y^{(0)} = y_0
\end{equation}
It has been proved that if $f(x,y)$ is bounded in some region about the point $(x_0, y_0)$ and if $f(x,y)$ satisfies the \textit{Lipschitz condition}, viz.,
\begin{equation*}
  |f(x,y)-f(x,\overline{y})| \leq K |y - \overline{y}|
\end{equation*}
for some constant $K$, then the sequence $y^{(n)}$ converges to the solution of the equation ~\ref{ivp}.

\subsection{Exercise}
\begin{enumerate}
\item Estimate the value of $y(0.2)$ using Taylor's series method from the IVP:  \\ $\displaystyle y'' -xy' -y = 0$ with $y(0) = 1$ and $y(0) =0$.

\item Find $y(0.1)$ correct to 4 decimal places using Taylor's series: $y'=x-y^2$, $y(0)=1$.

\item Solve the equation $y' = x+ y^2$, $y(0) = 1$ using Picard's method correct to $4$ decimal places and estimate the value of $y(0.2)$.
\end{enumerate}

@@

\subsection{Backward Euler's Method}
Using the backward difference $\left [ \displaystyle \frac{y_{n+1} -y_n}{h} = y'_{n+1} \right ]$; we obtain, Backward Euler's Method as follows:
\begin{equation}
  \label{eq:9}
  y_{n+1} = y_n + h\,f(x_{n+1}, y_{n+1}) \hspace{10mm} n=0,1,2,...
\end{equation}
We obtain $y_{n+1}$ in $f(x_{n+1}, y_{n+1})$ from the Euler's forward method. Actually, \\
$y^{1}_{n+1} = y_n + h\,f(x_n,y_n) \hspace{5mm} n=0,1,2,...$ and $y_{n+1} = y_n + h\,f(x_{n+1}, y^{1}_{n+1}) \hspace{10mm} n=0,1,2,...$

\subsection{Modified Euler's Method}
In the Modified Euler's Method the formula for iteration is
\begin{equation}
  \label{eq:9}
  \begin{gathered}[b]
    y^{1}_{n} = y_n + h\,f(x_n,y_n) \hspace{5mm} n=0,1,2,... \\[1mm]
     y_{n+1} = y_n + \frac{h}{2} \,\left [ f(x_n,y_n) + f(x_{n+1} + y^{1}_{n}) \right ]
  \end{gathered}
\end{equation}

\section{Runge-Kutta Methods}
Euler's method is less efficient in practical problems. It requires $h$ to be small for obtaining reasonable accuracy. To overcome this issue, Runge-Kutta methods are designed.
Let $k_0 =hf(x_0,y_0)$ and $k_1 = hf(x_1, y_0 + hf(x_0,y_0)$. The special case of second order Runge-Kutta Method gives $\displaystyle y_1 = y_0 + \frac{h}{2}(k_0 + k_1)$, which is equivalent to Modified Euler's Method. So, the iteration formula for the \textbf{second-order Runge-Kutta Method} is:
\begin{equation}
  \label{rungesecond}
  \begin{aligned}[b]
    k_n &= f(x_n,y_n) \\[1mm]
    k_{n+1} &= f(x_{n+1},y_n + hk_n) \\[1mm]
    y_{n+1} &= y_n + \frac{h}{2} \,\left [ k_n + k_{n+1} \right ]
  \end{aligned}
\end{equation}

\subsection{Fourth-Order Runge-Kutta Method}
The iteration formula for the Fourth-Order Runge-Kutta method is as follows:
\begin{equation}
  \label{rungefourth}
  \begin{aligned}[b]
    k_n &= f(x_n,y_n) \\[1mm]
    k_{n+1} &= f(x_n +(h/2) ,y_n + (h/2)k_n) \\[1mm]
    k_{n+2} &= f(x_n +(h/2) ,y_n + (h/2)k_{n+1}) \\[1mm]
    k_{n+3} &= hf(x_n + h ,y_n + hk_{n+2}) \\[1mm]
    y_{n+1} &= y_n + \frac{h}{6} \,\left [k_n + 2k_{n+1} + 2k_{n+2} + k_{n+3}\right ]
  \end{aligned}
\end{equation}

>>>>

\section{Taylor's series method}
We consider the IVP ~\ref{ivp}. If $y(x)$ is the exact solution of this IVP, then the Taylor's series for $y(x)$ around $x=x_0$ is given by
\begin{equation}
  \label{ivptay}
  y(x) = y_0 + (x-x_0)y'_0 + \frac{(x-x_0)^2}{2!}\, y''_0 + ...
\end{equation}
Differentiating ~\ref{ivp} successively, with respect to $x$, we obtain, $y'', y''', ...$. Since ~\ref{ivptay} is an infinite series, for $y(x_0)$ correct to $N$ decimal places we solve the following equation \\
$n^{th}$ term of Taylor series $\leq$ Upper limiting error  $\displaystyle \implies \frac{(x-x_0)^n}{n!} \, f^{(n)} (x_0) \leq \frac{1}{2} 10^{-N}$. \\[1mm]
If this holds then, the $n$ number of terms is sufficient otherwise, we need to include additional terms.
\begin{figure}[h]
  \centering
\includegraphics[width=13cm, height=10cm]{taylor_method_example.png}
\end{figure}

\section{Picard's Method of Successive Approximation}
Integrating the above differential equation ~\ref{ivp}, as follows, we obtain:
\begin{equation}
  \label{picard1}
  \begin{gathered}[b]
    \int_{x_0}^x \; y'\,dx = \int_{x_0}^x \; f(x,y)\,dx \\
      y = y_0 + \int_{x_0}^x \; f(x,y)\,dx
  \end{gathered}
\end{equation}
This equation in which the unknown function $y$ appears under the integral sign is called an \textit{integral equation}. Such equation can be solved by the method of successive approximations as follows:
\begin{equation}
  \label{picard2}
  y^{(1)} = y_0 + \int_{x_0}^x \; f(x,y_0)\,dx
\end{equation}
Proceeding in this way, we obtain:
\begin{equation}
  \label{picardit}
    y^{(n)} = y_0 + \int_{x_0}^x \; f(x,y^{(n-1)})\,dx \hspace{5mm} \text{with} \hspace{3mm}  y^{(0)} = y_0
\end{equation}
It has been proved that if $f(x,y)$ is bounded in some region about the point $(x_0, y_0)$ and if $f(x,y)$ satisfies the \textit{Lipschitz condition}, viz.,
\begin{equation*}
  |f(x,y)-f(x,\overline{y})| \leq K |y - \overline{y}|
\end{equation*}
for some constant $K$, then the sequence $y^{(n)}$ converges to the solution of the equation ~\ref{ivp}.
\begin{figure}[h]
  \centering
\includegraphics[width=13cm, height=10cm]{picard_method_example.png}
\end{figure}
\subsection{Exercise}
\begin{enumerate}
    \item Find by Taylor's series method:
\begin{enumerate}
    \item the value of $y(0.1)$ given that $y'=x-y^2$, $y(0)=1$.

    \item the value of $y(0)=1$ given that $\displaystyle \frac{dy}{dx} = 1 + xy$ correct up to four decimal places.
    \end{enumerate}
\item Use Picard's method to obtain:
\begin{enumerate}
    \item $y(0.1)$ and $y(0.2)$ given that $\displaystyle \frac{dy}{dx} = x + yx^4, \;\;\; y(0)=3$.
    \item $y(0.1)$ given that $\displaystyle \frac{dy}{dx} = \frac{y-x}{y+x}, \;\;\; y(0)=1$.
    \end{enumerate}
\end{enumerate}

@@

\subsection{Backward Euler's Method}
Using the backward difference $\left [ \displaystyle \frac{y_{n+1} -y_n}{h} = y'_{n+1} \right ]$; we obtain, Backward Euler's Method as follows:
\begin{equation}
  \label{eq:9}
  y_{n+1} = y_n + h\,f(x_{n+1}, y_{n+1}) \hspace{10mm} n=0,1,2,...
\end{equation}
We obtain $y_{n+1}$ in $f(x_{n+1}, y_{n+1})$ from the Euler's forward method. Actually, \\
$y^{1}_{n+1} = y_n + h\,f(x_n,y_n) \hspace{5mm} n=0,1,2,...$ and $y_{n+1} = y_n + h\,f(x_{n+1}, y^{1}_{n+1}) \hspace{10mm} n=0,1,2,...$
\begin{figure}[h]
  \centering
\includegraphics[width=13cm, height=10cm]{euler.png}
\end{figure}
\subsection{Modified Euler's Method}
In the Modified Euler's Method the formula for iteration is
\begin{equation}
  \label{eq:9}
  \begin{gathered}[b]
    y^{1}_{n} = y_n + h\,f(x_n,y_n) \hspace{5mm} n=0,1,2,... \\[1mm]
     y_{n+1} = y_n + \frac{h}{2} \,\left [ f(x_n,y_n) + f(x_{n+1} + y^{1}_{n}) \right ]
  \end{gathered}
\end{equation}
\subsection{Exercise}
\begin{enumerate}
    \item Use Euler's method to find $y(0.5)$ using $h=0.1$:
    \begin{multicols}{2}
        a). $\displaystyle \frac{dy}{dx}=\frac{3}{5}x^3y, \;\;\; y(0)=1$
        \columnbreak

        b). $\displaystyle \frac{dy}{dx}=1+y^2, \;\;\; y(0)=0$
    \end{multicols}

    \item Use Euler's modified method to find $y(0.2)$ and $y(0.4)$ using $h=0.2$:\\  $\displaystyle \frac{dy}{dx}=x+y, \;\;\; y(0)=0$.
    \end{enumerate}

\begin{figure}[h!]
  \centering
  \includegraphics[width=14cm, height=10cm]{modified.png} \\[2cm]

  \includegraphics[width=14cm, height=10cm]{rungee.png}
\end{figure}
\section{Runge-Kutta Methods}
Euler's method is less efficient in practical problems. It requires $h$ to be small for obtaining reasonable accuracy. To overcome this issue, Runge-Kutta methods are designed.
Let $k_0 =hf(x_0,y_0)$ and $k_1 = hf(x_1, y_0 + hf(x_0,y_0)$. The special case of second order Runge-Kutta Method gives $\displaystyle y_1 = y_0 + \frac{h}{2}(k_0 + k_1)$, which is equivalent to Modified Euler's Method. So, the iteration formula for the \textbf{second-order Runge-Kutta Method} is:
\begin{equation}
  \label{rungesecond}
  \begin{aligned}[b]
    k_n &= f(x_n,y_n) \\[1mm]
    k_{n+1} &= f(x_{n+1},y_n + hk_n) \\[1mm]
    y_{n+1} &= y_n + \frac{h}{2} \,\left [ k_n + k_{n+1} \right ]
  \end{aligned}
\end{equation}

\subsection{Fourth-Order Runge-Kutta Method}
The iteration formula for the Fourth-Order Runge-Kutta method is as follows:
\begin{equation}
  \label{rungefourth}
  \begin{aligned}[b]
    k_n &= f(x_n,y_n) \\[1mm]
    k_{n+1} &= f(x_n +(h/2) ,y_n + (h/2)k_n) \\[1mm]
    k_{n+2} &= f(x_n +(h/2) ,y_n + (h/2)k_{n+1}) \\[1mm]
    k_{n+3} &= hf(x_n + h ,y_n + hk_{n+2}) \\[1mm]
    y_{n+1} &= y_n + \frac{h}{6} \,\left [k_n + 2k_{n+1} + 2k_{n+2} + k_{n+3}\right ]
  \end{aligned}
\end{equation}
\subsection{Exercise}
\begin{enumerate}
    \item Use Runge-Kutta fourth order formula
    \begin{enumerate}
        \item to find $y(0.2)$ and $y(0.4)$ given $\displaystyle y'=\frac{y^2-x^2}{y^2+x^2}, \;\;\; y(0)=1$.

        \item to find $y(1.2)$ and $y(1.4)$ given $\displaystyle y'=\frac{3x+y}{x+2y}, \;\;\; y(1)=1$.

        \item to find $y(0.2)$ given $\displaystyle y'' +y=0, \;\;\; y(0)=1, \;\; y'(0)=0$.
    \end{enumerate}

  \item Given $\displaystyle \frac{dy}{dx}=y(1+x^2), \;\;\; y(0)=1$ find the values for $x=0.2, 0.4, 0.6$ using the Euler, the modified Euler, and the fourth order Runge-Kutta methods. Compare the results thus obtained.
  \item Solve the boundary value problems by finite difference method taking $y(0)=0, \;\; y(1)=1$ for $y=0.5, 0.25$:
    \hspace{3mm} a). $y''-y=0$ \hspace{7mm} b).$y''-x^2y'-e^xy=1/(1+x)$
\end{enumerate}

### 2025-07-06 sandwich_theorem/revised_sandesh_article.tex
chap 7

\documentclass[a4paper,twoside,12pt]{article}
\linespread{1}                     % For line spacing
\usepackage{amssymb}                 % For AMS symbol
\usepackage{amsmath,amsthm,latexsym}                 % This is useful for the matrix $\begin{bmatrix} a & b\\ 0 & c\end{bmatrix}$
\usepackage{verbatim}                % For comments in paragraph
\usepackage{graphicx}
\usepackage{epstopdf}
\usepackage{epsfig}              % For imbeding graph
\usepackage{color}                  % Color for the text
\usepackage[normal]{caption2}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{multicol}
\setlength{\columnsep}{0.8cm}
\usepackage{makeidx}
\usepackage[normal]{subfigure}
\usepackage{url}
\usepackage{lineno}
\usepackage{hyperref}
\usepackage{mathtools}
\usepackage{plain}
%%%%%%%%%%%%%
\textheight23cm%25.0cm
\textwidth16.3cm
\oddsidemargin-0.1cm
\evensidemargin0.1cm
%\topmargin-1cm
\headsep.3cm
%\usepackage{pstricks}
\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhf{}% Clear header/footer
\fancyhead[OC]{Sandesh Thakuri and Bishnu Hari Subedi}%Author on Odd page, Centred
\fancyhead[EC]{The Thakuri's Variant of Sandwich Theorem}% Title on Even page, Centred
\fancyfoot[C]{\thepage}%

@@

\title{\bfseries A Variant of the Sandwich theorem for two-sided limits}

@@

%---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------
\begin{document}
\linenumbers
\vspace{2mm}
{\Large
\begin{center}
\bf{\LARGE \bfseries A Variant of the Sandwich theorem for two-sided limits}
\end{center}}
\begin{center}
Sandesh Thakuri$^{1}$, Bishnu Hari Subedi$^{2}$
\end{center}

@@

\vspace{5mm}
\noindent
\textbf{Abstract:} {The criterion for the classical Sandwich theorem for two-sided limits is that the two-side limits must exists for the bounding functions. We show that this criterion can be relaxed. We show that it is sufficient for the existence of the left-hand limit for the lower bound function and the existence of right-hand limit for the upper bound function; and of-course they must be equal. This paper relaxes the criterion of the classical Sandwich theorem by replacing the two-sided limits with one-sided limits in the criterion and gives a new variant of the Sandwich theorem. While Rudin has given a proof of the Sandwich theorem for two sided limits and many has formulated the Sandwich theorem for the one-sided limits, these still don't relax the criterion of the Sandwich theorem for two-sided limits ~\cite{rudin},~\cite{thomas}. They have incorporated the one-sided limits for the Sandwich theorem for one-sided limits but has not relax the condition for the Sandwich theorem for two-sided limits as we have done.
\section{Introduction}
The Sandwich theorem is simple yet powerful tool in analysis to determine and to analyze the limit of a function at a given point. We can leverage the known limits to calculate the unknown limits. Suppose, we know the limits of \(g(x)\) and \(h(x)\) at \(x=c\) to be the same limit \(L\) and here \(f(x)\) happens to be sandwich between \(g(x)\) and \(h(x)\) in some neighborhood of \(c\). Then we can conclude the limit of \(f(x)\) at \(x=c\) as \(L\) by the Sandwich theorem for two-sided limits. The Sandwich theorem for two-sided limits is simply called the Sandwich theorem which is as follows.

@@

As an illustration, we know the limit of \(1/x\) is \(0\) as \(x
\to \infty\). This implies the limit of \(-1/x\) is also \(0\) as \(x \to \infty\). These are the known limits.\\
The value of sine function lies between \(-1\) and \(1\) so that \[\displaystyle -\frac{1}{x} \leq \frac{sinx}{x} \leq \frac{1}{x}\]
Now, \(\displaystyle \lim_{x \to \infty} -\frac{1}{x}=0\) and \(\displaystyle \lim_{x \to \infty} \frac{1}{x}=0\) so we can use the Sandwich theorem to conclude that \(\displaystyle \lim_{x \to \infty} \frac{sinx}{x}=0\).
\vspace{5mm}
\subsection{Limitation of the Sandwich Theorem}
The Sandwich theorem requires the existence of the two sided limits for the lower bound function $g(x)$ and the upper bound function $h(x)$. This is the limitation of the sandwich theorem. \\
We know \(\displaystyle - \left | x \right | \leq sinx \leq \left | x \right | \) which implies \[-\frac{\left | x \right |}{x} \leq \frac{sinx}{x} \leq \frac{\left | x \right |}{x}\]
But we cannot conclude the limit of \(\displaystyle \lim_{x \to 0} \frac{sinx}{x}\) using the Sandwich Theorem with the bounds above. It is so, because the \(\displaystyle \lim_{x \to 0} \frac{|x|}{x}\) does not exist. \\[2mm]
The \(\displaystyle \lim_{x \to 0} \frac{|x|}{x}\) does not exist because left hand limit and right hand limit of \(\displaystyle \frac{|x|}{x}\) at x=0 are different. They are different but they are finite.\\
\(\displaystyle \lim_{x \to 0^-} \frac{|x|}{x}=-1\) \hspace{5mm} and \hspace{5cm} \(\displaystyle \lim_{x \to 0^+} \frac{|x|}{x}=1\). \\[3mm]
We are not claiming that the limitation of the Sandwich theorem is not being able to determine the \(\displaystyle \lim_{x \to 0} \frac{sinx}{x}\). We are stating that the limitation of Sandwich theorem is its requirement of  the existence of the two sided limits for the bounding functions thus not being able to determine the above limit with the above bounds.
\vspace{5mm}
\section{The Thakuri's Variant of Sandwich Theorem for two sided limits}
\begin{theorem}[The Thakuri's Variant Sandwich Theorem]
Suppose $g(x) \leq f(x) \leq h(x)$ in some open interval containing $c$, expect possibly at $x=c$ itself. Suppose also that $$\lim_{x \to c^-} g(x) = \lim_{x \to c^+} h(x) =L  .$$ Then $\displaystyle \lim_{x \to c} f(x)=L$.
\end{theorem}
Now using this variant Sandwich theorem we can use the above bounds to conclude that \(\displaystyle \lim_{x \to 0} \frac{sinx}{x}=1\) as follows. \\[2mm]
 \(\displaystyle \lim_{x \to 0^-} -\frac{|x|}{x}=1\) \hspace{7cm}
 \(\displaystyle \lim_{x \to 0^+} \frac{|x|}{x}=1\) \\[1mm]
 Hence \(\displaystyle \lim_{x \to 0} \frac{sinx}{x}=1\).

 \subsection{Remarks on the our variant of Sandwich theorem}
 \begin{itemize}
 \item  While the \(\displaystyle \lim_{x \to 0} \frac{sinx}{x}\) can be determined by the sandwich theorem using the tighter-bound $cosx \leq sinx/x \leq 1$ ~\cite{boules}, the purpose of our variant of sandwich theorem is not to assert that this limit cannot be determined by the Sandwich theorem. This limit can be determined without the sandwich theorem as well.

 \item The purpose of our variant of sandwich theorem is to loosen the criterion of the sandwich theorem so that even for the bounds as illustrated above where two sided limits do not exits, it is applicable to use the sandwich theorem.

\item \textbf{The question of whether there exists any limit which can be determined by our variant but not by the classical sandwich theorem is another research problem that our research brings to the mathematics community.}
 \end{itemize}

\vspace{5mm}
 \subsection{Proof of the Variant of the Sandwich Theorem}

@@

\section{Previous Works on One-sided limits and Inequalities}
\subsection{Hardy's discussion on one-sided limits}
The proper use of one-sided limits in comparing the limits of two functions was done by Hardy in his book. There Hardy shows that inequalities between functions are preserved in the limit but the strictness of the inequalities is not preserved. That is, if $g(x) < f(x)$ for $x \in (c-\delta ,c )$, for some $\delta >0$ and $\lim_{x \to c^-} g(x) = L$, $\lim_{x \to c^+} f(x) =M$, then $L \leq M$ ~\cite{hardy}. But the converse is not true as Hardy shows, $L \leq M$ does not necessarily imply $g(x) < f(x)$. \\[2mm]
This result is crucial in handling the inequalities in limits. Hardy highlights that limits "smooth out" strict inequalities, converting $<$ or $>$ into $\leq$ or $\geq$ ~\cite{hardy}. This is crucial for setting inequalities in calculus (e.g., \textbf{sandwich theorem}) and understanding continuity and differentiability.

@@

\end{thm}
The proof of these one-sided theorems directly follows from the proof, given by Rudin, of the Sandwich theorem for two-sided limits ~\ref{two-side} ~\cite{rudin}.

\subsection{Distinction of our Variant from the other's one-sided limit variant}
\begin{itemize}
\item While the variant ~\ref{one-side} of Sandwich theorem incorporates one-one sided limits; this variant is for one-sided limits only. Our variant of Sandwich theorem not only incorporates the one-sided limits on the criterion and but is for two-sided limits.

\item The variant ~\ref{one-side} and other variant of the Sandwich theorem for the one-sided limits is analogous to the classical Sandwich theorem for two-side limits. Our variant is not analogous to the classical Sandwich theorem. These variants  is for the one-side limit. Our variant has left-hand limit and right-had limit only to loosen the criterion and it is not for the one-sided limits. It is for the two-sided limits.
\end{itemize}

\section{Significance of the Thakuri's Variant}
\begin{enumerate}
\item The Thakuri's version of Sandwich theorem for two-sided limits has made significance in relaxing the criterion of the classical Sandwich theorem for two sided limits. This version of Sandwich theorem allows us to use the even the bounds for which only one-sided limits exists.

\item Another significance of the Thakuri's version is it has brought another research problem to the mathematics community. The question of whether there exists any limit which can be determined by our variant but not by the classical sandwich theorem.

@@

\section{Conclusion}
\begin{itemize}
\item Even though the Thakuri's variant of Sandwich theorem seems obvious, its novelty is subtle and it still is a new variant which has the above significance.

\item For the common problems Thakuri's variant may not have a significant advantage but for some specific problem it can have a significant advantage. Besides having a new way to solve problem can give new perspective and new insights.

>>>>

\documentclass[a4paper,twoside,12pt]{article}
\linespread{1}                     % For line spacing
\usepackage{amssymb}                 % For AMS symbol
\usepackage{amsmath,amsthm,latexsym}                 % This is useful for the matrix $\begin{bmatrix} a & b\\ 0 & c\end{bmatrix}$
\usepackage{verbatim}                % For comments in paragraph
\usepackage{graphicx}
\usepackage{epstopdf}
\usepackage{epsfig}              % For imbeding graph
\usepackage{color}                  % Color for the text
\usepackage[normal]{caption2}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{multicol}
\setlength{\columnsep}{0.8cm}
\usepackage{makeidx}
\usepackage[normal]{subfigure}
\usepackage{url}
\usepackage{lineno}
\usepackage{hyperref}
\usepackage{mathtools}
\usepackage{plain}
%%%%%%%%%%%%%
\textheight23cm%25.0cm
\textwidth16.3cm
\oddsidemargin-0.1cm
\evensidemargin0.1cm
%\topmargin-1cm
\headsep.3cm
%\usepackage{pstricks}
\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhf{}% Clear header/footer
\fancyhead[OC]{Sandesh Thakuri and Bishnu Hari Subedi}%Author on Odd page, Centred
\fancyhead[EC]{The Thakuri's Extension of the Sandwich Theorem}% Title on Even page, Centred
\fancyfoot[C]{\thepage}%

@@

\title{\bfseries An Extension of the Sandwich theorem for two-sided limits}

@@

%---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------
\begin{document}
\linenumbers
\vspace{2mm}
{\Large
\begin{center}
\bf{\LARGE \bfseries An Extension of the Sandwich theorem for two-sided limits}
\end{center}}
\begin{center}
Sandesh Thakuri$^{1}$, Bishnu Hari Subedi$^{2}$
\end{center}

@@

\vspace{5mm}
\noindent
\textbf{Abstract:} {The criterion for the classical Sandwich theorem for two-sided limits is the existence of two-sided limit for the bounding functions. We show that this criterion can be relaxed. We prove that it is sufficient for the existence of the left-hand limit for the lower bound function and the existence of right-hand limit for the upper bound function; and of-course they must be equal. This paper relaxes the criterion of the classical Sandwich theorem by replacing the two-sided limits with one-sided limits in the criterion and gives an extension of the Sandwich theorem. While Rudin ~\cite{rudin} has given a proof of the Sandwich theorem for two sided limits and many has formulated the Sandwich theorem for the one-sided limits, these still don't relax the criterion of the Sandwich theorem for two-sided limits ~\cite{thomas}. They have incorporated the one-sided limits for the Sandwich theorem for one-sided limits but has not relax the condition for the Sandwich theorem for two-sided limits as we have done.
\section{Introduction}
The Sandwich theorem is simple yet powerful tool in analysis to determine and to analyze the limit of a function at a given point. We can leverage the known limits to calculate the unknown limits. Suppose, we know the limits of \(g(x)\) and \(h(x)\) at \(x=c\) to be the same limit \(L\) and here \(f(x)\) happens to be sandwich between \(g(x)\) and \(h(x)\) in some neighborhood of \(c\). Then we can conclude the limit of \(f(x)\) at \(x=c\) as \(L\) by the Sandwich theorem for two-sided limits. The Sandwich theorem for two-sided limits is simply called the Sandwich theorem which is as follows.

@@

As an illustration, we know the limit of \(1/x\) is \(0\) as \(x
\to \infty\). This implies the limit of \(-1/x\) is also \(0\) as \(x \to \infty\). These are the known limits.\\
The value of sine function lies between \(-1\) and \(1\) so that \[\displaystyle -\frac{1}{x} \leq \frac{sinx}{x} \leq \frac{1}{x}\]
Now, \(\displaystyle \lim_{x \to \infty} -\frac{1}{x}=0\) and \(\displaystyle \lim_{x \to \infty} \frac{1}{x}=0\) so we can use the Sandwich theorem to conclude that \(\displaystyle \lim_{x \to \infty} \frac{sinx}{x}=0\).
\vspace{5mm}
\subsection{Limitation of the Sandwich Theorem}
The Sandwich theorem requires the existence of the two sided limits for the lower bound function $g(x)$ and the upper bound function $h(x)$. This is the limitation of the sandwich theorem. \\
We know \(\displaystyle - \left | x \right | \leq sinx \leq \left | x \right | \) which implies \[-\frac{\left | x \right |}{x} \leq \frac{sinx}{x} \leq \frac{\left | x \right |}{x}\]
But we cannot conclude the limit of \(\displaystyle \lim_{x \to 0} \frac{sinx}{x}\) using the Sandwich Theorem with the bounds above. It is so, because the \(\displaystyle \lim_{x \to 0} \frac{|x|}{x}\) does not exist. \\[2mm]
The \(\displaystyle \lim_{x \to 0} \frac{|x|}{x}\) does not exist because left hand limit and right hand limit of \(\displaystyle \frac{|x|}{x}\) at x=0 are different. They are different but they are finite.\\
\(\displaystyle \lim_{x \to 0^-} \frac{|x|}{x}=-1\) \hspace{5mm} and \hspace{5cm} \(\displaystyle \lim_{x \to 0^+} \frac{|x|}{x}=1\). \\[3mm]
We are not claiming that the limitation of the Sandwich theorem is not being able to determine the \(\displaystyle \lim_{x \to 0} \frac{sinx}{x}\). We are stating that the limitation of Sandwich theorem is its requirement of  the existence of the two sided limits for the bounding functions thus not being able to determine the above limit with the above bounds.
\vspace{5mm}
\section{Thakuri's Extension of the Sandwich Theorem for two sided limits}
\begin{theorem}[Thakuri's Extension]
  \label{thakuri}
Suppose $g(x) \leq f(x) \leq h(x)$ in some open interval containing $c$, expect possibly at $x=c$ itself. Suppose also that $$\lim_{x \to c^-} g(x) = \lim_{x \to c^+} h(x) =L  .$$ Then $\displaystyle \lim_{x \to c} f(x)=L$.
\end{theorem}
Now using the Thakuri's extension of the Sandwich theorem we can use the above bounds to conclude that \(\displaystyle \lim_{x \to 0} \frac{sinx}{x}=1\) as follows. \\[2mm]
 \(\displaystyle \lim_{x \to 0^-} -\frac{|x|}{x}=1\) \hspace{7cm}
 \(\displaystyle \lim_{x \to 0^+} \frac{|x|}{x}=1\) \\[1mm]
 Hence \(\displaystyle \lim_{x \to 0} \frac{sinx}{x}=1\).

 \subsection{Remarks on the our extension of Sandwich theorem}
 \begin{itemize}
 \item  While the \(\displaystyle \lim_{x \to 0} \frac{sinx}{x}\) can be determined by the sandwich theorem using the tighter-bound $cosx \leq sinx/x \leq 1$ ~\cite{boules}, the purpose of our extension of the Sandwich theorem is not to assert that this limit cannot be determined by the Sandwich theorem. This limit can be determined without the Sandwich theorem as well.

 \item The purpose of our extension of the Sandwich theorem is to loosen the criterion of the sandwich theorem so that even for the bounds as illustrated above where two sided limits do not exits, it is applicable to use the sandwich theorem.

\item \textbf{The question of whether there exists any limit which can be determined by our extension but not by the classical sandwich theorem is another research problem that our research brings to the mathematics community.}
 \end{itemize}

\vspace{5mm}
 \subsection{Proof of Thakuri's Extension of the Sandwich Theorem}

@@

\section{Previous Works on One-sided limits and Inequalities}
\subsection{Hardy's discussion on one-sided limits}
The proper use of one-sided limits in comparing the limits of two functions was done by Hardy in his book ~\cite{hardy}. There Hardy shows that inequalities between functions are preserved in the limit but the strictness of the inequalities is not preserved. That is, if $g(x) < f(x)$ for $x \in (c-\delta ,c )$, for some $\delta >0$ and $\lim_{x \to c^-} g(x) = L$, $\lim_{x \to c^+} f(x) =M$, then $L \leq M$ ~\cite{hardy}. But the converse is not true as Hardy shows, $L \leq M$ does not necessarily imply $g(x) < f(x)$. \\[2mm]
This result is crucial in handling the inequalities in limits. Hardy highlights that limits "smooth out" strict inequalities, converting $<$ or $>$ into $\leq$ or $\geq$ ~\cite{hardy}. This is crucial for setting inequalities in calculus (e.g., \textbf{sandwich theorem}) and understanding continuity and differentiability.

@@

\end{thm}
The proof of these one-sided theorems directly follows from the proof, given by Rudin, of the Sandwich theorem ~\ref{two-side} for two-sided limits ~\cite{rudin}.

\subsection{Distinction of our extension from the other's:}
\begin{itemize}
\item While the variant, Theorem ~\ref{one-side} of Sandwich theorem incorporates one-one sided limits; this variant is for one-sided limits only. Our Theorem ~\ref{thakuri} not only incorporates the one-sided limits on the criterion and but is for two-sided limits.

\item The variant, Theorem ~\ref{one-side} and other variant of the Sandwich theorem for the one-sided limits is analogous to the classical Sandwich theorem for two-side limits. These variants are for the one-side limits. Our variant is for the two-sided limits. The left-hand limit and right-had limit in our variant is only to loosen the criterion of the Sandwich theorem.
\end{itemize}

\section{Significance of the Thakuri's Extension}
\begin{enumerate}
\item The Thakuri's extension of Sandwich theorem for two-sided limits has made significance in relaxing the criterion of the classical Sandwich theorem for two sided limits. This version of Sandwich theorem allows us to use the even the bounds for which only one-sided limits exists.

\item Another significance of the Thakuri's extension is, it has brought another research problem to the mathematics community. The question of whether there exists any limit which can be determined by our extension but not by the classical sandwich theorem.

@@

\section{Conclusion}
\begin{itemize}
\item Even though the Thakuri's extension of Sandwich theorem seems obvious, it is still is a new variant which the significance discussed above.

\item For the common problems Thakuri's extension may not have a significant advantage but for some specific problem it can have a significant advantage. Also, having a new way to solve problem can give new perspective and new insights.
